Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-09 3.Thesis/Writeup/2.Manuscript.tex
Last feedback before hand-in

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. \textbf{Therefore, algorithmic convergence should be monitored carefully, }although not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

@@

\textbf{{[}Include sub-questions and make structure of the paper clear!{]}} How can non-convergence be identified diagnostically? Are common MCMC non-convergence diagnostics appropriate for MICE? And if so, which threshold should be used to diagnose non-convergence? How many iterations are sufficient/needed to be able to diagnose non-convergence? Are the default number of iterations sufficient (i.e., 5 in mice, 10 in SPSS and Stata, 30 in mi)? How severe is it when the algorithm has not converged? And what are guidelines for practice? Can the parameter of interest \(Q\) be correct when the algorithm is not (yet) converged, and vice versa?

@@

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)). \textbf{In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle.} Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\). Ergo, non-convergence leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. \textbf{The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.}

@@

The aim of the simulation study is to determine the \textbf{impact of non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in iterative imputation algorithms. Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm.

\textbf{{[}REMOVE?{]} Several quantities of scientific interest \(Q\). Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{. And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice. That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).}

@@

\textbf{{[}REMOVE?{]} \(Q\)s are the targets that will be estimated in each simulation repetition, after first \emph{amputing} the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the \(Q\)s on the \(m\) completed datasets, and finally pooling the \(m\) estimates \(\hat{Q}_\ell\) to obtain a single \(\bar{Q}\) for each \(Q\). We try to predict the effect of non-convergence (i.e., the difference between \(Q\) and \(\bar{Q}\)) with the convergence diagnostics \(\widehat{R}\) and \(AC\), by applying these methods to several scalar summaries of the state space of the algorithm, \(\theta\)s.}

@@

In each simulation repetition, we \emph{ampute} the complete dataset five times to obtain five different incomplete datasets, \(y\)s. The ratio between \(y_{\rm obs}\) and \(y_{\rm mis}\) in each \(y\) depends on the missingness proportion: \(p_{\rm mis} =.05,.25,.5,.75,.95\). We ampute the complete data with the \texttt{mice} package \citep[function \texttt{mice::ampute()};][]{mice}. We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\).

Missing datapoints in each incomplete dataset \(y\) are imputed with the \texttt{mice()} function \citep{mice}. All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). Because \(m=5\), each run of the \texttt{mice()} function results in five sets of imputations: \(\dot y_{\rm imp, \ell}\), where \(m = 1, 2, ..., 5\). The number of iterations in each run of the algorithm varies between simulation conditions (\(T=1,2,...,100\)).

From the 5 imputations \(\dot y_{\rm imp, \ell}\) we obtain the chain means and chain variances for each variable \(j\). These will serve as scalar summary \(\theta\) to apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) on. We then use \texttt{mice::complete()} to combine the imputed data \(\dot y_{\rm imp, \ell}\) with the observed data \(y_{\rm obs}\), resulting in five completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}.

On the 5 completed datasets \{\(y_{\rm obs}, y_{imp, m}\)\}, we:

\begin{itemize}
\item
  Estimate \(\hat{Q}_\ell\)s for \(Q= \mu\) and \(Q=\sigma\) with \texttt{base::mean()}, and \texttt{stats::sd()} \citep{R}. We pool the \(\hat{Q}_\ell\)s with \texttt{base::mean()} to get \(\bar{Q}\)s.
\item
  Compute \(\lambda_{1, \ell}\), to use as \(\theta\) in the two convergence diagnostics. By definition, the first eigenvalue of the variance-covariance matrix, \(\lambda_{1, \ell}\), is equal to the variance of the principal component solution of each \{\(y_{\rm obs}, y_{imp, m}\)\}, which we calculate with \texttt{stats::princomp()} \citep{R}.
\item
  Perform multiple linear regression with \texttt{stats::lm()} \citep{R} to get the \(\hat{Q}_\ell\)s for \(Q=\beta\) and \(Q=r^2\). We pool the \(\hat{Q}_\ell\)s conform \citet{vink14} to get \(\bar{Q}\)s for the \(\beta\)s and \(r^2\) \textbf{{[}check finite pooling of \(r^2\), instead of \texttt{mice::pool.r.squared()}{]}}. Additionally, we use the \(\hat{Q}_\ell\)s of \(Q=\beta\) as \(\theta\) to apply the convergence diagnostics to.
\end{itemize}

We apply non-convergence diagnostics \(\widehat{R}\) and \(AC\) to each scalar summary \(\theta\). We calculate \(\widehat{R}\) by implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration.

@@

We assess the performance of the iterative imputation algorithm in each simulation condition by comparing \(\bar{Q}\)s with \(Q\)s. For each \(Q\), we calculate bias as \(\bar{Q} - Q\). For the regression coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage rate (CR). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(\rm SE_{\bar{Q}}\) is the square root of the pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we can conclude that \(\bar{Q}\) is a confidence-valid estimate of \(Q\). Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\). CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

@@

The estimated standard deviations not affected substantially by \textbf{early stopping} either. While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually \emph{less} severe than in other \textbf{iteration conditions}. Therefore, no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\). Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}=.05\). Conditions where \(p_{\rm mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion. Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is negligible. We therefore conclude that neither of the univariate \(Q\)s is affected substantially by non-convergence.

@@

The estimated coefficient of determination is clearly affected by early stopping. As expected, we observe worse performance in conditions with a lower number of iterations. The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in \(p_{\rm mis}\). \textbf{And so does} the number of iterations necessary to reach stable, approximately unbiased estimates. In conditions where \(p_{\rm mis}=.05\), estimates are unbiased after one iteration. We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively. The highest number of iterations necessary to reach approximate unbiasedness is seven. Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.25\). This implies that even for biased estimates, \(T\geq7\) would suffice to reach a stable solution.

@@

The \(\widehat{R}\)-values show a similar trend across iterations as we observed for chain means and chain variances. Only conditions with very high missingness proportions (\(p_{\rm mis}\geq.75\)) diverge from the earlier observations. The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations. The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\).

@@

@@

\hypertarget{conlusion}{%
\section{Conlusion}\label{conlusion}}

We arrive back at the original questions that we posed.

\begin{itemize}
\item
  Non-convergence can be identified with non-convergence diagnostics \(\widehat{R}\) and \(AC\). But common thresholds are too conservative. Why? Part of the distribution is already determined by the observed data. in regular MCMC the entire distribution is unknown (\textbf{priors??}). In iterative imputation algorithms we have the same characteristics as MCMC in the limit, but because we combine the known distribution with the unknown distribution, convergence is reached much sooner.
\item
  Inferences are unbiased much quicker than the algorithm has converged. All things univariate are converged almost intantly after initiating the iterative algorithm with a valid imputation model. But less quick with multivariate parameters.
\end{itemize}

In short, iterative imputation algorithms may yield approximately unbiased, confidence-valid estimates under different levels of induced non-convergence. Univariate estimates seem robust against terminating the algorithm early: There is no clear effect of the number of iterations on the bias in estimated means and standard deviations. The bias in these \(Q\)s was only impacted by the missingness proporion \(p_{\rm mis}\), with higher \(p_{\rm mis}\) resulting in more bias. Bias in multivariate estimates depended on both \(p_{\rm mis}\) and the number of iterations. However, the only effect of the number of iterations was observed when the algorithm was terminated at seven iterations or earlier. This suggests that continued iteration beyond seven iterations does not yield better estimates. Yet, the non-convergence diagnostics \(\widehat{R}\) and \(AC\) indicated that the iterative imputation algorithm did not reach a stable state before twenty to thirty iterations. \textbf{{[}MERGE or remove{]} \(\widehat{R}\) and autocorrelation indicate that there are signs of non-convergence in the algorithm up-to at least twenty iterations, while unbiased, confidence-valid estimates may be obtained with as little as one iteration. This observation is in agreement with }one of {[}still add the others!{]}** the simulation hypotheses: \(\widehat{R}\) over-estimates the severity of non-convergence when applied to MI procedures.**

Over-estimation of the signs of non-convergence may be due to the methods (and their thresholds) or due to the \(Q\)s (descriptive and multivariate linear regression, not higher dimensional/more complex \(Q\)s). More complicated \(Q\)s (e.g., higher-order effects or variance components) might show bias, under- or over-coverage at higher \(T\). Which might be why \(\widehat{R}\) and \(AC\) identify non-convergence.

Some other interesting findings are \textbf{{[}REPHRASE{]}}

\begin{itemize}
\item
  Is there more bias in conditions where non-convergence was induced? (YES for multivar., NOPE for univariates).
\item
  Is the point of \emph{sufficient} convergence (at which the \(\bar{Q}\)s are unbiased and confidence-valid estimates of the \(Q\)s) the same point as identified by the non-convergence diagnostics? (NOPE, they are too conservative).
\item
  Is one of the thetas better? (YES, while it doesn't really matter as long as it's multivariate, \(\lambda_{1, \ell}\) is model-independent and therefore preferred).
\item
  What should the thresholds be? (much HIGHER than recommended for \(\widehat{R}\), difficult to say for \(AC\)).
\item
  And the default nr of iterations? (MAYBE increase to 10 to detect significant \(AC\) and be sure not to miss the dip in \(\widehat{R}\)?).
\end{itemize}

@@

After imputation it is always wise to take the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  First, check traceplots for signs of pathological non-convergence. Adjust imputation model if necessary.
\item
  Monitor non-convergence diagnostics \(\widehat{R}\) and \(AC\). Do not use the traditional calculation for \(\widehat{R}\), or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\). Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
\item
  Track \(\theta\)s over iterations, e.g., the novel \(\theta\) that is `substantive model-independent'.
\item
  Track your own scalar summary of interest (see e.g., \citet{buur18}). Compute \(\widehat{R}\) and autocorrelation values for this scalar summary.
\item
  Use the threshold \(\widehat{R}>1.2\) for mixing, and assess stationarity by plotting the autocorrelation over iterations. Keep iterating until the threshold for mixing is reached, and until a reasonably decreasing asympototic \(AC\) value is obtained. At that point, inferences are unlikely to improve by continued iteration.
\end{enumerate}

\hypertarget{recommendations-for-future-research}{%
\subsection{Recommendations for future research}\label{recommendations-for-future-research}}

Much remains unknown. about non-convergence in iterative imputation algorithms. Further research is needed to investigate the performance of iterative imputation algorithms under clear violation of convergence, e.g.~dependency between predictors (predictors with very high correlations). For example:

\begin{itemize}
\item
  Induce non-convergence with mis-specified imputation model.
\item
  Introduce more difficult missingness problems, i.e., M(N)AR instead of MCAR. That is, proper performance under a `missing completely at random' missingness mechanism is necessary to demonstrate the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However, results may not be extrapolated to other missingness mechanisms.
\item
  Imputation technique with questionable convergence properties, e.g., PMM.
\item
  Apply recommendations to empirical data and write a vignette for applied researchers.
\item
  Implement recommendations in imputation software, e.g., evaluation environment for applied researchers.
\item
  Develop an additional convergence diagnostic unique for iterative imputation: \(AC\) across imputations, instead of iterations.
\end{itemize}

In short, we have shown that an iterative imputation algorithm can yield correct outcomes, even when a converged state has not yet formally been reached. Any further iterations would then burn computational resources without improving the statistical inferences. To conclude, in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity. \textbf{{[}Our study found that - in the cases considered - inferential validity was achieved after 5-10 iterations, much earlier than indicated by the \(\widehat{R}\) and AC diagnostics. Of course, it never hurts to iterate longer, but such calculations hardly bring added value.{]}}

>>>>

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied \citep{buur18}. There is no scientific consensus on how to evaluate the convergence of imputation algorithms \citep{taka17}. Moreover, the behavoir of such algorithms under certain default imputation models (e.g., `predictive mean matching') is an entirely open question \citep{murr18}. Therefore, algorithmic convergence should be monitored carefully, \textbf{but this is not straightforward}. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence \citep{gelm13}. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostics may only identify signs of \emph{non}-convergence \citep{hoff09}. Several of such non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms.

@@

@@

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren \citeyearpar[\(\S\) 6.5.2]{buur18}. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain: scalar summary \(\theta\) plotted against the iteration number \(t\). The \(\theta\) in this example is the chain mean of a variable that we will call \(j\) (i.e., the average of the imputed values for \(j\) in each imputation \(\dot y_{\rm imp, \ell}\)). In the typical convergence scenario, the imputation chains intermingle nicely and there is little to no trending. In the non-convergence scenario, there is a lot of trending and some chains do not intermingle. Importantly, the chain means at the last iteration (the imputed value per imputation \(\ell\)) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a factor two larger than those of the typically converged algorithm. Non-convergence thus influences the mean of \(j\) in \(\dot y_{\rm imp, \ell}\). This difference (presumably) translates to the \(m\) completed data estimates \(\hat{\mu}_{j,\ell}\), and consequently to the pooled estimate \(\bar{\mu}_j\). Ergo, non-convergence leads to biased estimates. This shows the importance of reaching converged states in iterative imputation algorithms.

@@

\textbf{{[}MERGE with sections above{]}} Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain \(\theta\). \(\theta\)s are scalar summaries of the state space of the algorithm at a specific iteration. The default scalar summaries in \texttt{mice} are chain means and chain variances. Additionally, researchers may ``monitor some statistic of scientific interest'' \citep[\(\S\) 6.5.2]{buur18}. As \citet{buur18} describes, researchers can specify their quantity of scientific interest \(Q\) (e.g., a regression coefficient) as scalar summary \(\theta\). Such a user-defined scalar summary, however, may be somewhat advanced for empirical researchers. And moreover, it is not universal to all complete data problems. Focusing on the convergence of outcome parameters may influence the iterative imputation procedure in the sense that the model of evaluation favors the model of interest. The downside to these two approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. Therefore, we propose a \(\theta\) that summarizes the multivariate state space of the algorithm, but is independent from the model of scientific interest. We propose \(\lambda_{1, \ell}\) as such a scalar summary. We define \(\lambda_{1, \ell}\) as the first eigenvalue of the variance-covariance matrix of the completed data. Let \(\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}\) be the eigenvalues of \(\Sigma_\ell\) in each of the \(m\) completed datasets \{\(y_{\rm obs}, \dot y_{\rm imp, \ell}\)\}. \(\lambda_{1, \ell}\) is measure that summarizes the covariances in the completed datasets. The first eigenvalue has the appealing property that is not dependent on the substantive model of interest. Eigenvalue decomposition is inspired by \citet{mack03}.\textbf{{[}ADD definition variance in first PCA component is equal to the first eigenvalue of the variance-covariance matrix{]}}

@@

The aim of the simulation study is to determine the \textbf{impact of non-convergence {[}``early stopping'' en ``proportion of complete cases''{]}} in iterative imputation algorithms. Specifically, we are interested in the bias and confidence-validity of estimates obtained from incomplete data with the MICE algortihm. \textbf{{[}REMOVE?{]} Subsequently we will evaluate how well \(\widehat{R}\) and \(AC\) perform in identifying the effects of non-convergence }{[}and performance of different thetas{]}\textbf{. And finally, we will formulate an informed advice on the requirements to safely conclude sufficient convergence in practice. That is, we conclude that convergence is sufficient when each estimate \(\bar{Q}\) is an unbiased and confidence-valid estimate of the corresponding estimand \(Q\).}

@@

@@

The complete dataset is \emph{amputed} in each simulation repetition to create five incomplete datasets \(y\) with different missingness proportions \(p_{\rm mis}\). The \(p_{\rm mis}\) varies between 5 and 95\%, in incremental steps (\(p_{\rm mis}\) = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a `missing completely at random' missingness mechanism \citep{rubin87}. I.e., the probability to be missing is the same for all \(N \times k\) cells in \(y\), conditional on the missingness proportion \(p_{\rm mis}\). This is performed with the `mice' package \citep[function \texttt{mice::ampute()};][]{mice}.

The second set of simulation conditions is the number of iterations in the imputation algorithm. We vary the number of iterations between one and one hundred (\(T = 1,2,.,100\)). All imputation procedures are performed with Bayesian linear regression imputation, and five imputation chains (\(m=5\)). We perform this with the \texttt{mice()} function \citep{mice}. The result is five sets of imputations per simulation condition, \(y_{\rm imp, \ell}\).

Subsequently, we obtain the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\} by combining the imputed data with the observed data \citep[function \texttt{mice::complete()};][]{mice}. Univariate quantities of scientific interest \(Q\) are estimated directly from the completed data \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. Estimates of multivariate \(Q\)s are obtained after employing multiple linear regression \citep[function \texttt{stats::lm()},][]{R}. From the estimates in each imputation, \(\hat{Q}_\ell\), we obtain the pooled estimate \(\bar{Q}\). For univariate estimates we pool \(\hat{Q}_\ell\) as average. Multivariate \(\hat{Q}_\ell\) are pooled conform \citet{vink14}.

\hypertarget{methods.}{%
\subsubsection{Methods.}\label{methods.}}

In total, we consider eight methods to identify non-convergence. Namely, we apply the two non-convergence diagnostics \(\widehat{R}\) and \(AC\) on four different \(\theta\)s. Univariate \(\theta\)s are obtained from \(y_{\rm imp, \ell}\). The multivariate \(\theta\) \(\lambda_{1,\ell}\) is obtained from \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. The other multivariate \(\theta\) is equal to the estimated regression coefficient \(\hat{Q}_\ell\) in \{\(y_{\rm obs}, y_{\rm imp, \ell}\)\}. We apply the two non-convergence diagnostics to each \(\theta\) by implementing \citet{veht19} 's recommended \(\widehat{R}\) and we compute \(AC\) manually.

@@

As recommended by \citet{buur18} we evaluate the performance of the imputation algorithm with bias, coverage rate (CR), and confidence interval width (CIW). Bias is calculated for each \(Q\). CR and CIW only for \(Q = \beta\).
We calculate bias as \(\bar{Q} - Q\). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(m-1)} \times \rm SE_{\bar{Q}},\]
where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(m-1\) degrees of freedom, and \(\rm SE_{\bar{Q}}\) is the square root of the pooled variance estimate. If we obtain nominal coverage (CR = 95\%), we can conclude that \(\bar{Q}\) is a confidence-valid estimate of \(Q\). Finally, we inspect CI width (CIW): the difference between the lower and upper bound of the 95\% confidence interval around \(\bar{Q}\). CIW is of interest because it is a measure of efficiency. Under nominal coverage, short CIs are preferred, since wider CIs indicate lower statistical power.

@@

The estimated standard deviations not affected substantially by \textbf{early stopping} either. While stable estimates are obtained in conditions with two iterations, the bias in the first iteration is actually \emph{less} severe than in other \textbf{iteration conditions}. Therefore, no iteration at all is needed to obtain approximately unbiased estimates for \(Q=\sigma\). Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}=.05\). Conditions where \(p_{\rm mis}\geq.25\) are impacted by non-convergence in order of increasing missingness proportion. Similar to the bias for \(Q=\mu\) the magnitude of the bias in \(Q=\sigma\) is negligible. We therefore conclude that neither one of the univariate \(Q\)s is affected substantially by non-convergence.

@@

The estimated coefficient of determination is clearly affected by early stopping. As expected, we observe worse performance in conditions with a lower number of iterations. The magnitude of the bias, however, depends on the missingness proportion---following the incremental order in \(p_{\rm mis}\). \textbf{And so does} the number of iterations necessary to reach stable, approximately unbiased estimates. In conditions where \(p_{\rm mis}=.05\), estimates are unbiased after one iteration. We observe approximately unbiased estimates when \(T\geq 2,4, \text{and }5\), in conditions where \(p_{\rm mis} = .25, .50, \text{and }.75\), respectively \textbf{{[}only report the worst?{]}}. The highest number of iterations necessary to reach approximate unbiasedness is seven. Completely unbiased estimates are only obtained in conditions where \(p_{\rm mis}\leq.25\). This implies that even for biased estimates, \(T\geq7\) would suffice to reach a stable solution.

@@

The \(\widehat{R}\)-values show a similar trend across iterations as we observed for chain means and chain variances. Only conditions with very high missingness proportions (\(p_{\rm mis}\geq.75\)) diverge from the earlier observations. The \(\widehat{R}\)-values in these conditions do not taper off after 30 to 40 iterations, but rather between 40 to 50 iterations. The highest number of iterations at which non-convergence is still diagnosed according to the thresholds are: 11 for \(\widehat{R}>1.2\), 23 for \(\widehat{R}>1.1\), and 100 for \(\widehat{R}>1.01\) \textbf{{[}only report the worst?{]}}.

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

We have shown that non-convergence in iteratiive imputation algorithms leads to biased, invalid estimates of multivariate quantities of scientific interest \(\theta\). Since the current practice of visually inspecting univariate \(\theta\)s does not suffice, we applied non-convergence diagnostics \(\widehat{R}\) and \(AC\) to both univariate and multivariate \(\theta\)s. Signs of non-convergence due to early stopping are identified with each of the methods---i.e., by applying \(\widehat{R}\) and \(AC\) to any \(\theta\). Bias due to high missingness proportions, is only identified with multivariate \(\theta\)s. Under moderate missingness (\(p_{\rm mis}\leq.50\)) \(\widehat{R}\)-values decreased substantially until 30 to 40 iterations, which implies that mixing in the algorithm improved with each additional iteration. \(AC\)-values only improved until about six iterations, suggesting minimal improvement in trending is obtained beyond \(T=6\). Since we obtained approximately unbiased, confidence valid estimates after at most seven iterations, we conclude that the \(\widehat{R}\) threshold 1.2 is the most appropriate diagnostic cut-off. The threshold to diagnose non-stationarity with \(AC\) does not seem to apply. The main finding of this study is that valid inferences may be obtained much quicker than approximate algorithmic convergence is reached. Under the current specifications, univariate \(Q\)s did not require algorithmic convergence at all. They are unbiased almost instantly after the algorithm is initiated. Approximately unbiased, confidence-valid estimates of multivariate \(Q\)s were obtained after a maximum of seven iterations. Continued iterations beyond \(T = 7\) did not yield better estimates.

\hypertarget{implications}{%
\subsection{Implications}\label{implications}}

Iterative imputation algorithms such as MICE have been known to yield valid results under severe missingness. With this study, we have shown that valid inferences can also be obtained under a combination of severe missingness and early stopping. Based on our results, the traditional threshold of \(\widehat{R}> 1.2\) would be appropriate for diagnosing non-convergence in iterative imputation algorithms. This is not in line with the recent recommendations by veht19 to lower the threshold to 1.01. The discrepancy between our conclusion and \citet{veht19}'s may be explained by the nature of iterative imputation algorithms. In the limit, iterative imputation algorithms have the same character as other MCM algorithms, but in imputation procedures a part of the distribution is already determined by the observed data, whereas the entire distribution is unknown in many other MCMC applications. Because we combine a known distribution with an unknown distribution, valid estimates may be reached much sooner. Convergence may therefore be diagnosed at a less stringent threshold. \textbf{{[}HERE were the two asterixes{]}}

The threshold to diagnose non-convergence with \(AC\) does not seem appropriate at all in iterative imputation algorithms. A better diagnostic cut-off would be the number of iterations at which an additional iteration does not substantially decrease \(AC\). In practice, this implies that the default number of iterations in MICE is not sufficient. \(AC\) can only be computed if \(T\geq3\), while the current default in the \texttt{mice} package is five \citep{mice}. Identifying a decrease in the \(AC\)-values across iterations (an `elbow') requires more than three observations. We therefore suggest to increase the default number of iterations in MICE to 10. The computational cost of 5 additional iterations has become less of a burden since MICE was introduced in \citeyearpar{mice}. Moreover, an increased number of iterations would grant the opportunity to exclude the first iterations from evaluation with \(\widehat{R}\). Excluding \(T\leq3\) provides empirical researchers with a less ambiguous heuristic to diagnose non-convergence, because we expect the initial `dip' in \(\widehat{R}\)-values to disappear.

@@

Based on our results, we formulated several recommendations for empirical researchers who employ iterative imputation to draw inferences from incomplete data. We suggest that non-convergence may still be evaluated visually, but in addition to inspecting univariate summaries of the state space of the algorithm (\(\theta\)s, e.g., chain means), multivariate \(\theta\)s should be considered. We propose the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  First, check traceplots of the default \(\theta\)s (chain means and chain variances) for signs of pathological non-convergence. Adjust the imputation model if necessary.
\item
  Track multivariate \(\theta\)s over iterations, e.g., the novel \(\theta\) that is scientific model-independent. Or specify your own scalar summary of interest (see e.g., \citet{buur18}). Monitor these \(\theta\)s across iterations through visual inspection, or using a non-convergence diagnostic.
\item
  Compute non-convergence diagnostics \(\widehat{R}\) and \(AC\). Do not use the traditional calculation for \(\widehat{R}\) \citep{gelm92}, or the \texttt{R} function \texttt{stats::acf()} to compute \(AC\) \citep{R}. Instead, calculate \(\widehat{R}\) conform \citet{veht19} and compute autocorrelations manually (see e.g., \href{https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/1.SimulationStudy}{github.com/gerkovink/shinyMice}).
\item
  Use the threshold \(\widehat{R}>1.2\) to diagnose non-mixing, and assess stationarity by plotting the autocorrelation over iterations. Keep iterating until the threshold for mixing is overcome, and until reasonably decreasing asympototic \(AC\)-values are obtained. At that point, inferences are unlikely to improve by continued iteration.
\item
  Profit.
\end{enumerate}

\hypertarget{limitations}{%
\subsection{Limitations}\label{limitations}}

Much remains unknown about non-convergence in iterative imputation algorithms. Even though we have demonstrated the appropriateness of \(\widehat{R}\) and \(AC\) as non-convergence diagnostics in this study, results may not extrapolate to other situations. The current simulation conditions were restricted to a single missingness mechanism. Proper performance under a `missing completely at random' (MCAR) mechanism is a necessary condition for any missing data method. It doesn't, however, guarantee equal performance under different missingness mechanisms. Future research should determine the performance of \(\widehat{R}\) and \(AC\) under `missing at random' and `missing not at random' mechanisms.
Another parameter to consider in future research is the choice of the imputation method. As \citet{murr18} concluded, the behaviour of algorithms such as MICE under certain default imputation models is still an open research question. We have investigated the behaviour of MICE under only one type of model (Bayesian linear regression). Different imputations models might converge more poorly. Moreover, the imputation model may even be mis-specified un purpose to induce different levels of non-convergence. Qualitatively, we have shown that \(\widehat{R}\) and \(AC\) are appropriate under clear violation of convergence. Quantitatively, we have only demonstrated that they can identify signs of non-convergence under MCAR, when the imputation model is a correctly specified Bayesian linear regression model. \newline \newline
